A sustainable approach to tourist signage on heritage trails\footnote{This research has been supported by the Ministry of University and Research under the "Underlandscape" project (2020428LS8)}

### 2023-11-07 paper.tex
Revisione introduzione e conclusioni e altro

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}
\usepackage{float}
\usepackage[caption = false]{subfig}
\usepackage{tablefootnote}
\usepackage{multibib}
\usepackage{listings}
\usepackage{xcolor}

@@

Leisure activities play are relevant in bringing profit to niche businesses. This creates a reinforcing effect that follows the negative dynamics seen above: a leisure activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact. Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.

@@

This document addresses the specific aspect of heritage trail signage, and describes a solution based on QR code tags. The design gathers suggestions from several disciplines to take into account the many aspects that contribute to sustainability. The result is a practical signage technology that is adaptable to other environments sharing defined traits.

@@

Such a solution is severely limited by the power supply dependency. A radio transmitter is a rather power-consuming device, in the range of Watts. Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.

@@

The NFC technology is currently very diffused, although reading an NFC tag as text requires the installation of an appropriate application. Once the application is running the operation simply consists in approaching the smartphone to the tag: the tag content is transferred to the {\em user's} device as a chink of text. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection.

Another solution in the family of passive devices is the QR tag (QR stands for Quick-Response). Such a technology does not require radio communication but uses the smartphone camera to acquire a graphical code that is translated into text. The capacity of a QR tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR tag roughly equals that of an NFC tag. A QR tag is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.

@@

A preliminary check verifies compliance with the Universal Design Principles \cite{udi97a}:

@@

Such minimal solutions (see figure \ref{fig:qrnfccoin}) compare well with other more complex ones that make use of the user's smartphone. There is a trade-off concerning capacity, but in many circumstances, a capacity of 2-300 words is sufficient to convey the description of the site or provide directions for the visit. If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:

\begin{itemize} 
\item does not require any power supply,
\item has a limited impact on the landscape,
\item does not entice theft,
\item has a negligible cost,
\item is durable,
\item produces a limited quantity of waste when disposed of,
\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet.
\end{itemize}

@@

For a signage application, a QR tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalism and misuse.

In conclusion, we have reasons to select a QR-tag-based solution as a good candidate for smartphone-based signage. We now consider how it copes with the limits of a traditional board-based approach (as listed in table \ref{sec:signage}):

@@

\section{Materials and Methods}

@@

\subsection{The natural and cultural resources of \emph{Casoli} and its vicinity}

@@

\subsection{An integrated view of tourism in the \emph{Casoli} region}

@@

The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like the \emph{Orrido di Botri} with the \emph{Canyon Park\footnote{\url{ https://www.canyonpark.it/}}}, and other service companies and associations for experiential tourism.

In reason of a whole insight of the mentioned touristic area, a socio-cultural approach is needed, by considering also tourism flows. In this perspective, it must be considered that there isn’t touristic information at the administrative level of \emph{Casoli}, so the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT (National Statistical Institute).
 The demographic trend (Figure 4, a) enlightens a relevant aspect of \emph{Bagni di Lucca} society and economic aspects of its touristic ecosystem (Figure 4, b).

@@

\emph{Bagni di Lucca} stretches on 164.70 $km^2$ area with a population density \footnote{Index referring to the number of people living in a territory per square kilometer. It is the rate between the annual number of residents and the surface area. In our case and for 2022, the density is 5593 people over 164.70 $km^2$.} of 3395 inhabitants per square kilometer and a job occupancy rate of 33.53\%. Regarding human growth at the territorial level, it is relevant to consider the \emph{Bagni di Lucca} population trend during the 2001-2022 period. The curve shown by figure \ref{fig:population} registers a constant and slight decrease of one thousand people in a twenty-year time frame (2001: 6556 people; 2022: 5593 people).

@@

The Tourism Density Index \footnote{The Tourism Density Index shows the tourism concentration in the higher touristic season. It is the rate between the number of tourists and the surface area.} is 106.41 tourists per $km^2$ in 2022. The arrivals exhibit a drop from 2019 (from  4339 in 2017 to 2363 in 2021), and touristic presences have suffered a slight decrease during the same period, as shown in figure \ref{fig:tourismflows}, while a further increase has been registered in 2022 (5.876 arrivals).

@@

\begin{table}
    \footnotesize
    \setstretch{1,2}
    \centering
    \begin{tabular}{l||l|l} \hline 
        {\bf Region} & {\bf Bagni di Lucca} & {\bf Lucca} \\ \hline
        Resident density (residents/{$km^2$}) & 33.95 & 215.71 \\
        Touristic density (visitors/{$km^2$}) & 106.41 & 1866.96 \\
        Average stay  (days) &  2.98 &  3.36 \\
        Accommodation density (accommodations/{$km^2$}) & 10.14 & 19.30 \\
        Coexistence Index (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
         \hline
    \end{tabular}
    \caption{Tourist indexes in \emph{Bagni di Lucca} in 2022\tablefootnote{Each index gives a different type of touristic information about \emph{Bagni di Lucca} Municipality: the touristic density represents the number of tourists per $km^2$; the Average stay index is the average number of nights spent in town by tourists; the Accommodation Density Index is related to the number of touristic presences on the number of beds occupied by tourists all over the year; the coexistence index shows the relevance of foreign tourists per 100 Italian tourists.}}
    \label{tab:tourism}
\end{table}

@@

\subsubsection{Our activity}

@@

\subsubsection{Sites and trail’s state of art}

@@

\subsection{Implementing a QR-based signage}

@@

We aim to define a design approach that will enhance the tourist experience of an inland area. The approach is conceptual and therefore applicable to a wide range of use cases. First, the requirements for sustainability in design are defined, taking into account the three pillars of sustainability: social, environmental, and economic.

To put our approach into practice we used a small-scale use case: a signage system addressing the guidance of visitors in an inland area. The area has a cultural and environmental heritage to be protected and valorized at the same time.

Our design started by defining a profile for the area according to the \emph{three pillars}. The environmental aspect is characterized by natural formations and human artifacts tracing back to the Paleolithic Era. As for tourism-related economics, currently the dominant component in the area, we noticed that the location selected for our use case exhibits unexpressed potentialities although the area surrounding the location has a recognized touristic vocation. From the social point of view, we observed that the area is exposed to depopulation due to the shortage of productive activities.

One conclusion after the above analysis is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable. Regarding the potential reasons for a visit, we found a not easily accessible cave inhabited since pre-history, a village with historical buildings, and nearby architecture dating back to the Middle Ages.

We investigated a technique for guiding the visitor and telling the history of the place. The fragile economy limits financial investment, while the natural heritage requires non-intrusive techniques.

@@

The preparation of the signage required a solid knowledge of the area that was gained by carrying out archaeological surveys and consulting historical documents and cartography. The evaluation of touristic vocation was based on the available statistical data and person-to-person informal interviews. Two on-site surveys provided further practical indications and the opportunity to place the QR tags.

On the technical side, the proposed signage is a benchmark difficult to improve in terms of environmental impact and cost.

@@

In areas like our use case, where the cultural heritage extends across scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging. In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos. Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot. The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.

Collecting statistics about user perception is not a primary interest for us at this stage. However, we provided two ways to collect such data. One is a 5-questions poll on the web pages linked to each tag. The user is asked to agreement a level of agreement with the following sentences:
\begin {itemize}
\item the QR tag was useful
\item a plaque would be preferable
\item I scanned other QR codes in the vicinity
\item I will come back to visit the sites I did not visit today
\item I am satisfied about my visit
\end{itemize}
In addition, the hits on the pages linked to the QR-codes were counted. Both ways proved to be ineffective. In the case of the form, we collected one positive feedback that has no statistical significance. The hit count was polluted by spurious access to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
%Its {\em performance} is currently difficult to evaluate. The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons, one being that the QR tag cannot record or transmit read events. Another is that the user may be unable or unwilling to visit the URL in the tag. Finally, the URL can be visited by means different from the QR reading. For such reasons, despite we acknowledging its importance for the stakeholders, we do not discuss such an index.

@@

Geo-itinerary

@@

Based on qualitative and quantitative analysis \cite{loz09} carried out in the social and touristic context of Bagni di Lucca, including the Casoli village, it can be argued that the area under investigation has unique features that need to be capitalized on to enhance its tourism potential. Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for potential tourists practicing heterogeneous touristic circuits in individual and combined ways. Our study aims to involve stakeholders in sustainable planning for new heritage itineraries, despite accessibility limitations on site.

>>>>

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}
\usepackage{float}
\usepackage[caption = false]{subfig}
%\usepackage{tablefootnote}
\usepackage{multibib}
\usepackage{listings}
\usepackage{xcolor}

@@

Leisure and cultural activities, like tourism, bring profit to niche businesses. This creates a reinforcing effect that follows the negative dynamics seen above: a activity that is neutral in itself (e.g., skying) can have a significant impact due to induced effects (lifts) that strengthen the success of the activity (easier to practice) bringing more investment and impact. Unlike the beneficial effects, which are usually well localized, the negative impact has a very broad spectrum that includes environmental, social, and economic sectors, as discussed by Patthey in \cite{pat08a} relating to outdoor winter sports.

@@

This paper illustrates a concrete exercise in designing for sustainability. We introduce an instance of cultural sustainability, an inland region with a legacy to defend, and design a concrete tool to address the issue. To reach fully sustainable result, we need to practice from the early design stages an holistic approach to the environmental, social and financial facets of the problem.

The concrete aspect we address is a signage system, and the solution we propose is based on QR code tags. The result is applicable or adaptable to other environments sharing defined traits.

The next section illustrates signage as a topic of interest. The discussion targets low cost, low impact technologies that are appropriate for outdoor signage of natural or cultural heritage, and finally indicates QR code tags as a consistent technology. The section features a survey of relevant literature.

Section \ref{sec:methods} provides the multidisciplinary framework. Section \ref{sec:legacy} illustrates the legacy emphasizing historical facts and locations. Section \ref{sec:tourism} describes the social and touristic context of the target location and provides the guidelines for integrating the initiative in the socio-political fabric. Such guidelines are used to define an agenda, in section \ref{sec:agenda}, and the section is concluded by a reasoned description of the area, highlighting the points of interest and their reachability \ref{sec:soa}.

Section \ref{sec:implementation} describes the concrete implementation that follows the defined requirements. The description of the process to produce the printed tags starting from editable content on a geographic database allows to reproduce and reuse the solution in other contexts.

Section \ref{sec:results} merges the various aspects explored in the previous sections to provide a methodological view of the process, focusing on the interplay between the fourth pillars of sustainability.

The conclusive section summarizes the results, the missing parts, and the potential of the solution.

@@

Such a solution is severely limited by the dependency from a power supply. A radio transmitter is a rather power-consuming device, in the range of Watts. Even if intermittent, the operation of a battery-operated networking device cannot be guaranteed for long periods and the host organization should consider power harvesting, which negatively contributes to the economic cost and environmental footprint of the device.

@@

The NFC technology is available without installing ad-hoc applications. Its operation consists in approaching the smartphone to the tag, and the user is advertised that a content is available. The smartphone can read it aloud to compensate for the user's inability or translate to deal with linguistic issues. Such capabilities do not need an Internet connection but currently depend on ad-hoc applications.

Another solution in the family of passive devices is the QR tag (QR stands for Quick-Response). Such a technology does not require radio communication but uses the smartphone camera to acquire text encoded in a graphical code. The capacity of a QR tag depends on the number of dots in the image, which in turn depends on the size of the code and the smartphone camera characteristics. We may consider that the capacity of a QR tag roughly equals that of an NFC tag. A QR tag is larger than an NFC tag of comparable capacity, and manufacturing requires a printer.

@@

A preliminary check verifies compliance of the passive technologies described above with the Universal Design Principles \cite{udi97a}:

@@

Compared with other smartphone-based technologies, there is a trade-off concerning capacity. However, a signage applications a capacity of 2-300 words is sufficient to provide directions for the visit and convey text-based educational contents. If capacity limits are not an issue, a solution based on NFC or QR tags exhibits several advantages:

\begin{itemize} 
\item does not require power supply,
\item has a limited impact on the landscape,
\item does not entice theft,
\item has a negligible cost,
\item is durable,
\item produces a limited quantity of waste when disposed of,
\item content can be stored for later usage; for instance, to visit a URL once the user reaches a zone covered by the Internet,
\item is inclusive regarding user inabilities and nationality,
\item requires basic technical skills from the user.
\end{itemize}

@@

For a signage application, a QR tag is preferable because a noticeable size is needed. In addition, keeping the tag out of reach prevents vandalism and misuse.  Therefore we have reasons to select a QR-tag-based solution for smartphone-based signage.

Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging. In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos. Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot. The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.

We now examine how such technology copes with the limits of a traditional board-based approach (as listed in table \ref{sec:signage}):

@@

\section{Materials and Methods \label{sec:methods}}

@@

\subsection{The natural and cultural assets of \emph{Casoli} and its vicinity \label{sec:legacy}}

@@

\subsection{An integrated view of tourism in the \emph{Casoli} region \label{sec:tourism}}

@@

The strategic position of \emph{Bagni di Lucca} is one of the strengths of its tourism context, together with the presence of such attractive elements like the \emph{Orrido di Botri} \citeWeb{botri} with the \emph{Canyon Park} \citeWeb{canyonpark}, and other service companies and associations for experiential tourism.

In reason of a whole insight of the mentioned touristic area, a socio-cultural approach is needed, by considering also tourism flows. In this perspective, it must be considered that there isn’t touristic information at the administrative level of \emph{Casoli}, so the tourism context analysis concerns the Municipality area of \emph{Bagni di Lucca}, as the first territorial context providing tourist data registered by ISTAT (National Statistical Institute) \citeWeb{istat}.
 The demographic trend (Figure 4, a) enlightens a relevant aspect of \emph{Bagni di Lucca} society and economic aspects of its touristic ecosystem (Figure 4, b).

@@

\emph{Bagni di Lucca} stretches on 164.70 $km^2$ area with a population density of 33.95 inhabitants per square kilometer and a job occupancy rate of 33.53\%. Regarding human growth at the territorial level, it is relevant to consider the \emph{Bagni di Lucca} population trend during the 2001-2022 period. The curve shown by figure \ref{fig:population} registers a constant and slight decrease of one thousand people in a twenty-year time frame (2001: 6556 people; 2022: 5593 people).

@@

The Tourism Density Index is the rate between the number of tourists and the surface area in the higher touristic season. It is 106.41 tourists per $km^2$ in 2022. The arrivals exhibit a drop from 2019 (from  4339 in 2017 to 2363 in 2021), and touristic presences have suffered a slight decrease during the same period, as shown in figure \ref{fig:tourismflows}, while a further increase has been registered in 2022 (5.876 arrivals).

@@

\begin{table}
    \footnotesize
    \setstretch{1,2}
    \centering
    \begin{tabular}{l||l|l} \hline 
        {\bf Region} & {\bf Bagni di Lucca} & {\bf Lucca} \\ \hline
        Resident density (residents/{$km^2$}) & 33.95 & 215.71 \\
        Touristic density (visitors/{$km^2$}) & 106.41 & 1866.96 \\
        Average stay  (days) &  2.98 &  3.36 \\
        Accommodation density (accommodations/{$km^2$}) & 10.14 & 19.30 \\
        Coexistence (100 $\times$ Foreign/Italian) & 103.74\% &  69.87\% \\ 
         \hline
    \end{tabular}
    \caption{Tourist indexes in \emph{Bagni di Lucca}
%. The {\em touristic density} is the number of tourists in a year per $km^2$. The {\em average stay} is the average number of nights spent in town by tourists. The {\em accommodation density} is the rate between the number of touristic presences and the number of beds occupied by tourists in a year. The {\em Coexistence} is the percent rate between foreign and Italian tourists.
}
    \label{tab:tourism}
\end{table}

@@

\subsubsection{Our activity \label{sec:agenda}}

@@

\subsubsection{Sites and trail’s state of art \ref{sec:soa}}

@@

\subsection{Implementing a QR-based signage \label{sec:implementation}}

@@

We aim to enhance the tourist experience in an inland area located around the village of Casoli. Our goal is tightly related with the fourth pillar of sustainability: there are natural and cultural assets that need to be nurtured or at least preserved. We concentrate on a signage system, the realization of which is finalized to reaching such cultural result.

To avoid the over-tourism pitfall, we need to take into account the other three pillars: social, environmental, and economic. To this end we analyze the social fabric that will host our initiative, the environment where signage will be deployed, and the economic framework that will support its implementation. From this study we derive the requirements for our signage system: low environmental impact, open to participation of the residents, easy to use from the visitors, and economically sustainable with a minimal touristic flow.

The approach we applied to reach such conclusions is summarized below: it is conceptual and therefore applicable to a wide range of use cases, and merely consists in designing the system with sustainability in mind from the early stages.

Our design started by defining a profile for the area according to the \emph{three pillars}. The environmental aspect is characterized by natural formations and human artifacts tracing back to the Paleolithic Era. As for tourism-related economics, a qualitative and quantitative analysis \cite{loz09} reveals unique features that need to be capitalized on to enhance the tourism potential. Given the touristic trends and the local vocation, we conclude that Casoli is a small destination for visitors practicing heterogeneous touristic circuits in individual and combined ways. Cultural sustainability mandates the involvement of local stakeholders in planning new heritage itineraries and supporting initiatives, including signage.

An outcome of the analysis of the social fabric is that, in order to preserve the native social and economic framework, mass tourism is not a candidate target, while a motivated, non-casual visitor is preferable. Regarding the potential reasons for a visit, we found a cave inhabited since pre-history but reachable with some effort, a village with historical buildings, and nearby architecture dating back to the Middle Ages. Such targets offer a variety of access ways, from those requiring performance to the relaxing ones. For this reason we opted for not suggesting a geo-itineray, allowing the user to follow personal inclinations. In this way we expect, justified by the results of \cite{sol18a}, that the educational impact of the signage is maximized.

We investigated a technique for guiding the visitor and telling the history of the place. The small number of expected visitors limits financial investment, while the natural heritage requires non-intrusive techniques.

@@

Previous investigations, including archaeological surveys and access to historical documents and cartography, allowed to design the contents to be delivered through signage. Two on-site surveys provided further practical indications and the opportunity to place the QR tags.

The main result from our activity is the real scale deployment of the signage, a non trivial exercise in holistic sustainable design.

As a side result we have a signage technology based on QR code tags, a benchmark difficult to improve in terms of environmental impact and cost.

%Its use is especially appropriate when the cultural heritage extends across a scarcely accessible land, preserving and accessing sites located on private property or in hazardous areas can be challenging. In these cases, QR tags could improve the visitors' experience by providing, besides textual data, the Web URLs for videos, aerial shots, and photos. Multimedia resources allow visitors to immerse themselves in the area's history and gain an experiential understanding, even if they don't have access to a specific spot. The resulting widespread knowledge can be a beneficial tourist-economic engine that encourages investments in cultural heritage policies, thus achieving the goal of protecting and conserving abandoned sites.

@@

Collecting statistics about user perception is not a primary interest for us at this stage: the initiative started as an exercise and was not adequately advertised to the potential users. However, we provided two ways to collect data as part of the exercise.

One is a poll in five points on the web pages linked to each tag. The user is asked to quantify an agreement (0-do not agree, 5-strongly agree) with four sentences, and finally provide an answer:
\begin{itemize}
\item the QR tag was useful
\item a plaque would be preferable
\item I will come back to visit the sites I did not visit today
\item I am satisfied about my visit
\item I scanned other QR codes in the vicinity
\end{itemize}

The other is an hits count on the pages linked to the QR-codes.

Both of them proved to be ineffective. We collected only one form with a positive feedback but with no statistical significance. The hit count was polluted by spurious accesses to the page (for instance for maintenance, or reaching the page from the map) and by the fact that scanning the tag was not necessarily linked to a hit.
%Its {\em performance} is currently difficult to evaluate. The recording of the number of times a given tag is scanned, in theory, a significant performance index, is unreliable for technical reasons, one being that the QR tag cannot record or transmit read events. Another is that the user may be unable or unwilling to visit the URL in the tag. Finally, the URL can be visited by means different from the QR reading. For such reasons, despite we acknowledging its importance for the stakeholders, we do not discuss such an index.

@@

@@
